---
title: "immunoCluster"
author: "James Opzoomer, Kevin Blighe, Jessica Timms"
date: "`r Sys.Date()`"
package: "`r packageVersion('immunoCluster')`"
output:
  github_document:
    toc: true
    toc_depth: 2
fig_width: 7
always_allow_html: true
---


**NOTE: THIS PACKAGE IS STILL UNDER DEVELOPMENT AND SO SOME OF THE FUNCTIONALITY IS NOT FULLY TESTED**

## 1. Introduction to immunoCluster

The immunoCluster package provides a complete toolkit to carry out immune profiling from both liquid and imaging high-dimensional mass (CyTOF) and flow cytometry. The package features standardized data infrastructure, making use of the SingleCellExperiment class object, interactive visualization tools and convenient implementations for popular dimenionality reduction and clustering algorithms designed for a non-specialist. To learn using immunoCluster we have provided walkthrough below exploring some of the packages basic functionality.   

## 2. Main functionalities of immunoCluster

The basic pipeline of immunoCluster analysis involves:

* Creating file and marker labels to generate the experimental design

* Importing the .fcs data into the SingelCellExperiment object

* Exploratory analysis of the data

* Dimensionality reduction 

* Clustering and biological interpretation of the clusters

* Differential abundace and differential expression analysis

```{r, echo = FALSE, message = FALSE}
  library(knitr)
  library(kableExtra)
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE)
```

# Installation

## 1. Install the package from github

Install the development version of the package, installation will usually take a few minutes, depending on the number of dependencies that have been installed on your pc. Install immunoCluster from github here:

```{r getPackageDevel, eval = FALSE}
  # Install devtools for github installation if not present
  require(devtools)
  
  # Install the ImmunoCluster package
  devtools::install_github("kordastilab/ImmunoCluster")
  
  # May need to run this line if an R build version error while installing appears
  # Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")
```

## 2. Load the package and the dependancies into the R session


Load necessary packages:
```{r Load, eval=TRUE, results="hide"}
  library(immunoCluster)

  # Load dependencies
  require(flowCore)
  library(stringr)
  library(cowplot)
  library(RColorBrewer)
  library(Biobase)
  library(ConsensusClusterPlus)
  library(FlowSOM)
  library(limma)
  library(ggrepel)
  library(reshape2)
  library(Rtsne)
  library(dplyr)
  library(pheatmap)
```

# Walkthrough: Immune profiling PBMC from a two condition experimental design

## 1. Introduction

In this walkthrough we will learn to perform immune profiling and examine the from changes in immune populations from human PBMC. We will start with pre-gated .fcs data files. The samples were taken from patients following hematopoetic reconsitution in leukemia patients following BMT. The dataset represents 15 individuals, sampled at multiple time points after BMT, for a total of 28 samples. Of these patients, a small subset suffered from graft versus host disease (GvHD, n = 3), whereas most other patients did not experience such complications. 

We have performed gating cleanup on a subset of the dataset which can be dowloaded from zenodo, along with all the other data required for the walkthough.  [Insert zenodo button].

The data was originally reported in [Comprehensive Immune Monitoring of Clinical Trials to Advance Human Immunotherapy](https://www.sciencedirect.com/science/article/pii/S2211124719308228). The full raw data (not used here, but which was used in the immunoCluster paper) is available [here](http://flowrepository.org/id/FR-FCM-Z244).


## 2. Download and import data 

You can download all the .fcs files from zenodo here here, or alternatively you can download the SingleCellExperiment object directly into memory and skip the import process just below.

```{r download fcs, eval=FALSE}
# Insert zenodo link

```

The experimental design informs the what metadata we will need to incorporate into the SingleCellExperiment object, here we have created one table (from an excel file), that contains five columns, and one row for each .fcs file. We will include the .fcs filename, the GvHD "condition" of the patient,  patient_id, a unique sample_id, and the day_id representing the day following BMT that the sample was taken. (Link to excel file).

```{r sample setup, eval=TRUE}
# Excel file can be read in using this line
#

# Insert zenodo RDS link to replace local
sample_metadata_df = readRDS(file = "../github_readme/rds_objects/sample_metadata.rds")

head(sample_metadata_df)
```

We also include a table that represents a row for every channel in the original fcs file. The columns represent the original name if the channel, the name we want to rename it in antigen. We then optionally include columns of binary 1 and 0s that allow us to group markers into relevant sets, here for clustering or differential expression analysis. Any marker that is not included in either group will be discarded. This table must be an exact match to the order that the markers are represnted in the raw .fcs files. (Link to excel file).

```{r panel setup, eval=TRUE}
# Excel file can be read in using this line
#

# Insert zenodo RDS link to replace local
panel_metadata_df = readRDS(file = "../github_readme/rds_objects/panel_metadata.rds")

head(panel_metadata_df)

# Create marker grouping designations
clustering_markers = panel_metadata_df$antigen[panel_metadata_df$clustering ==1]
activation_markers = panel_metadata_df$antigen[panel_metadata_df$activation == 1]
col_names = panel_metadata_df$antigen[panel_metadata_df$clustering == 1 | panel_metadata_df$activation == 1 ]

# Select columns to dicard from fcs file
discard = setdiff(panel_metadata_df$antigen, c(clustering_markers, activation_markers))
queries = str_detect(panel_metadata_df$antigen, paste(discard, collapse = "|"))

if(length(discard) > 0){colsDiscard = panel_metadata_df$metal[queries]
}else{colsDiscard = ''}

```

We next load the .fcs data and the metadata into the SingleCellExperiment object. Here the number of rows in the metadata must match the number of .fcs files to be analysed. We have the option to specify the scaling transformation and the cofactor. We have used asinh with cofactor 5. It is possible to downsaple from each file or by a specified metadata grouping, although here we have not downsampled. The columns are also renamed as specified in the panel_metadata table and columns not needed are discarded.

```{r import data, eval=FALSE}

# Collect all .fcs files from a directory
filelist = list.files(
      path = "../github_readme/fcs_data",
    pattern = "*.fcs|*.FCS",
    full.names = TRUE)

# Select which metadata to include
metadata = data.frame(
      file = filelist, 
      group = sample_metadata_df$sample_id,
      condition = sample_metadata_df$condition,
      patient_id = sample_metadata_df$patient_id,
      day_id = sample_metadata_df$day_id,
      row.names = filelist,
      stringsAsFactors = FALSE)

# Create SCE object with CyTOF data inside
require(SingleCellExperiment)
  sce_gvhd = processFCS(
    files = filelist,
    metadata = metadata,
    transformation = TRUE, # Transform data
    filter = FALSE, 
    transFun = function (x) asinh(x), # Using asinh function (cofactor 5 is used by default)
    downsampleVar = NULL, 
    downsample = NULL, # Downsample to n cells per downsample_grouping
    downsample_grouping = NULL, # Downsample to downsample cells by the specified metadata slot
    newColnames = col_names, # Rename columns from panel metadata 
    colsDiscard = colsDiscard) # Discard columns not selected for downstream analysis

sce_gvhd
```

Using the code chunk below you can download the imported r data object with the metatdat incorporated directly into memory to skip the import steps and save time:

```{r download rds, eval=TRUE}
# Insert zenodo RDS link
sce_gvhd = readRDS(file = "../github_readme/rds_objects/sce_gvhd.rds")

sce_gvhd
```

## 2. Data Exploration

We provide several functions to examine globla patterns within the data, they can be used to get an idea of the influence of conditions/treatment or batch effects within the data. The mdsplot() function will generate a multi-dimensional scaling (MDS) plot on median expresion values for each channel. MDS plots represent an unsupervised way to indicate similarities between samples at a global level berfore more in depth analsysi. In the example here we can see that there is no observabe high-level difference between the two GvHD condition samples, however there does appear to be a weak separation between samples collected at D30 and D90 post BMT. Here and throughout the pipeline, colkey can be used to specify plotting colors for graphing conditions. 

```{r mds_plot, fig.width=11, fig.height=3.75, eval=TRUE}
mds1 = mdsplot(sce_gvhd, 
               feature = "condition", 
               colkey = c(None = 'royalblue', GvHD = 'red2'))

mds2 = mdsplot(sce_gvhd, 
               feature = "day_id", 
               colkey = c(D30  = 'darkorange1', D90 = 'darkgreen'))

plot_grid(mds1, mds2,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 20)
```

Further to the mds plots,the medianHeatmap() function will generate a heatmap with median marker expression across all (default) or a selection (with markers = ). The heatmap is clustered over rows and columns. The heatmap can be generated on the asinh tranformed values (default) or 0->1 scaled (with scale_01 = T). The heatmap should provide insight as to which markers are highly expressed and wether thier marker expression profile is assosciated with a metadata condition. Highly expressed markers will have a greater influence on clustering and this will help understand clustering results generated downstream.

```{r sample heatmap, eval=TRUE}
feature = medianHeatmap(sce_gvhd, grouping = "group", feature = "condition", feature_cols = c('royalblue', 'red2'))
```

## 3. Dimensionality reduction and clustering

immunoCluster supports multiple different dimensionality reduction algorithms and has UMAP and tSNE functionality implemented below with the ability to downsample before running the algorithms. We run both UMAP and tSNE on selected lineage markers, downsampling to 1000 cells per sample to reduce runtime. The dimensionality reductions are stored in the reducedDimanmes slot and any other dimensionality reduction, like PCA, can also be stored in the SCE object in parallel.

```{r run UMAP, message=FALSE, eval=TRUE}
require(umap)
# Run UMAP and store in sce object
sce_gvhd = performUMAP(sce_gvhd, downsample = 1000, useMarkers = clustering_markers)

# Run tSNE and store in sce object
sce_gvhd = performTSNE(sce_gvhd, downsample = 1000, useMarkers = clustering_markers)

sce_gvhd
```

```{r, fig.width=14, fig.height=10, eval=TRUE}
expression_markers =  c('CD3', 'CD4', 'CD8a', 'CD11b', 'CD19', 'CD56')

exp_plot_umap = markerExpression(sce_gvhd,
    markers = expression_markers,
    reducedDim = 'UMAP',
    title = 'Marker exprssion in UMAP space',
    nrow = 1, ncol = 6,
    pointSize = 0.05,
    legendKeyHeight = 1.0,
    legendLabSize = 14,
    stripLabSize = 20,
    axisLabSize = 18,
    titleLabSize = 24,
    captionLabSize = 22)

exp_plot_tsne = markerExpression(sce_gvhd,
    markers = expression_markers,
    reducedDim = 'Rtsne',
    title = 'Marker exprssion in tSNE space',
    nrow = 1, ncol = 6,
    pointSize = 0.05,
    legendKeyHeight = 1.0,
    legendLabSize = 14,
    stripLabSize = 20,
    axisLabSize = 18,
    titleLabSize = 24,
    captionLabSize = 22)

plot_grid(exp_plot_umap, exp_plot_tsne,
    labels = c('A','B'),
    nrow = 2, align = "l", label_size = 24)
```

Run consensus method of FLowSOM and ConsensusClusterPlus on all cells using a selection of markers, with a final desired cluster number of uo to 10. 

```{r run clustering, message=TRUE, eval=TRUE}
sce_gvhd = runFlowSOM(sce_gvhd, k = 10, markers = clustering_markers)
```

```{r viz flowSOM, fig.width=11, fig.height=6, eval=TRUE}
fsom_k10 = metadataPlot(sce_gvhd,
    colby = 'flowsom_cc_k10',
    title = 'flowSOM k=10',
    legendPosition = 'top',
    legendLabSize = 16,
    axisLabSize = 16,
    titleLabSize = 16,
    subtitleLabSize = 16,
    captionLabSize = 16)

fsom_clusters = unique(sce_gvhd@metadata$flowsom_cc_k10)

fsom_abundance = plotAbundance(sce_gvhd,
              clusters = fsom_clusters,
              clusterAssign = 'flowsom_cc_k10',
              feature = NULL,
              legendPosition = 'none', 
              stripLabSize = 10,
              axisLabSize = 14,
              titleLabSize = 12,
              subtitleLabSize = 10,
              captionLabSize = 18)


plot_grid(fsom_k10, fsom_abundance,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 20)
```

```{r diagnostic heatmap, eval=TRUE}
# With feature
feature = medianHeatmap(sce_gvhd, grouping = "flowsom_cc_k10", feature = NULL, scale_01 = T)
```

```{r enriched markers, eval=TRUE}
library(scran)
library(tibble)

cytof_markers = findMarkers(assay(sce_gvhd), sce_gvhd@metadata$flowsom_cc_k10, test.type="wilcox", direction="up",lfc=1.5, pval.type="some")

cluster_markers = NULL
top_n = 3
for (i in 1:length(cytof_markers)) {
  
  deg_markers = cytof_markers[[i]][1:top_n,1:2]
  
  deg_markers = rownames_to_column(as.data.frame(deg_markers), var = "marker")
  
  markers = data.frame(cluster = rep(i, top_n), deg_markers)
  
  cluster_markers = rbind(cluster_markers, markers)
  
}

# Top3 DEG markers per cluster
cluster_markers
```

```{r merge clusters, fig.width=11, fig.height=6, eval=TRUE}
# Assignments
original_cluster_ident = as.character(rep(1:10))
new_cluster_ident = c("Monocytes", "cDC", "NKT Cells", "CD4+ T Cells", "B Cells","gd T Cells", "CD8+ T Cells", "Basophils", "CD4+ Treg", "NK")

# Set the new annotations
sce_gvhd = setClusterIdents(sce_gvhd, orig.ident = original_cluster_ident , new.ident = new_cluster_ident, clustering = 'flowsom_cc_k10')

# Vizualise in UMAP space
cell_annotation = metadataPlot(sce_gvhd,
    colby = 'cell_annotation',
    title = 'Annotated clusters',
    legendLabSize = 8,
    axisLabSize = 20,
    titleLabSize = 20,
    subtitleLabSize = 16,
    captionLabSize = 16)

annotated_clusters = unique(sce_gvhd@metadata$flowsom_cc_k10)

annotated_abundance = plotAbundance(sce_gvhd,
              clusters = annotated_clusters,
              clusterAssign = 'cell_annotation',
              feature = NULL,
              legendPosition = 'none', 
              stripLabSize = 10,
              axisLabSize = 14,
              titleLabSize = 12,
              subtitleLabSize = 10,
              captionLabSize = 18)

plot_grid(cell_annotation, annotated_abundance, 
    labels = c('A','B'),
    ncol = 2, nrow = 1, align = "l", label_size = 20)
```
```{r bar plot, eval=TRUE}
bar_plot = plotAbundance(sce_gvhd,
              graph_type = 'bar',
              clusters = annotated_clusters,
              clusterAssign = 'cell_annotation',
              feature = 'condition',
              legendLabSize = 8,
              stripLabSize = 22,
              axisLabSize = 22,
              titleLabSize = 22,
              subtitleLabSize = 18,
              captionLabSize = 18)

bar_plot
```

```{r final heatmap, eval=TRUE}
# With feature
feature = medianHeatmap(sce_gvhd, grouping = "cell_annotation", feature = NULL, scale_01 = T, heat_bar = "Greys")
```



```{r cell annotations, fig.width=11, fig.height=8, eval=TRUE}
cell_annotation_dif = metadataPlot(sce_gvhd,
    colby = 'condition',
    title = 'UMAP labelled by GvHD status',
    colkey = c(None = 'royalblue', GvHD = 'red2'),
    legendPosition = 'top',
    legendLabSize = 16,
    axisLabSize = 20,
    titleLabSize = 20,
    subtitleLabSize = 16,
    captionLabSize = 16)

cell_annotation_dif = cell_annotation_dif + facet_wrap(~condition)

abundance_dif = plotAbundance(sce_gvhd,
              clusters = annotated_clusters,
              clusterAssign = 'cell_annotation',
              feature = 'condition',
              colkey = c(None = 'royalblue', GvHD = 'red2'),
              legendPosition = 'none', 
              stripLabSize = 10,
              axisLabSize = 14,
              titleLabSize = 12,
              subtitleLabSize = 10,
              captionLabSize = 18)

plot_grid(cell_annotation_dif, abundance_dif,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 20)
```

```{r abundance test, eval=TRUE}
 library(broom)
# This need the individual values
# This also need the right contrasts
porportions = diffClust(sce_gvhd, 
                        clustering = 'cell_annotation')

porportions[,c(1,14:17)]
```


```{r marker test, eval=TRUE}
# Need to catch errors - cluster not there causes crash
marker_test = diffExpression(sce_gvhd,
                assay = 'scaled',
                grouping = 'group', # The condensing function
                feature = 'condition', # The contrast
                clusterAssign = 'cell_annotation')

sig_markers = marker_test[which(marker_test$p.value < 0.05),]

sig_markers[c(1,12:14),c(13:19)]
```

```{r activation markers, eval=TRUE}

# Difference in activation markers per cluster
markerExpressionPerSample(sce_gvhd,
    caption = '',
    clusters = c("CD8+ T Cells", "cDC"),
    feature = 'condition',
    clusterAssign = 'cell_annotation',
    markers = activation_markers,
    colkey = c('royalblue', 'red2'),
    title = '',
    stripLabSize = 12,
    axisLabSize = 10,
    titleLabSize = 18,
    subtitleLabSize = 10,
    captionLabSize = 10)
```

```{r diffcyt, eval=TRUE}

library(diffcyt)



```

```{r subsetSCE, eval=TRUE}

# SubsetSCE function

```


# Session info
```{r session info, eval=TRUE}
sessionInfo()
```

# Contact

For any queries relating to software:

* James W Opzoomer (james.opzoomer@kcl.ac.uk)
