---
title: "immunoCluster"
author: "James Opzoomer, Kevin Blighe, Jessica Timms"
date: "`r Sys.Date()`"
package: "`r packageVersion('immunoCluster')`"
output:
  github_document:
    toc: true
    toc_depth: 2
fig_width: 7
always_allow_html: true
---




## 1. Introduction to immunoCluster

The ImmunoCluster package provides a complete toolkit for carry out immune profiling from High-dimensional mass (CyTOF) and flow cytometry. The package features standardized data infrastructure, using the SingleCellExperiment class object, interactive visualization tools and convenient implementations for popular algorithms designed for a non-specialist. Provided below is a simple walthrough exploring some of the packages functionality. 

## 2. Main functionalities of immunoCluster

**NOTE: THIS PACKAGE IS STILL UNDER DEVELOPMENT AND SO SOME OF THE FUNCTIONALITY IS NOT FULLY TESTED**

```{r, echo = FALSE, message = FALSE}
  library(knitr)
  library(kableExtra)
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE)
```

# Installation

## 1. Install the package from github

Install the development version of the package:

```{r getPackageDevel, eval = FALSE}
  # Install devtools for github installation if not present
  require(devtools)
  
  # Install the ImmunoCluster package
  devtools::install_github("kordastilab/ImmunoCluster")
```

## 2. Load the package and the dependancies into the R session

```{r Load, eval=FALSE}
  library(immunnoCluster)

  # Load dependencies
  require(flowCore)
  library(stringr)
  library(cowplot)
  library(RColorBrewer)
  library(Biobase)
  library(ConsensusClusterPlus)
  library(FlowSOM)
  library(limma)
  library(ggrepel)
  library(reshape2)
  library(Rtsne)
  library(dplyr)
  library(pheatmap)
```

# Walkthrough: Immune profiling PBMC from simple experimental setup

## 1. Introduction

In this walkthrough we will start with FCS data files to examine changes in immune populations from PBMC. The samples were taken from patients following hematopoetic reconsitution in leukemia patients following BMT. The dataset represents 15 individuals, sampled at multiple time points after BMT, for a total of 28 samples. Of these patients, a small subset suffered from graft versus host disease (GvHD, n = 3), whereas most other patients did not experience such complications. 

The data was originally reported in [Comprehensive Immune Monitoring of Clinical Trials to Advance Human Immunotherapy](https://www.sciencedirect.com/science/article/pii/S2211124719308228). 

The raw data is available [here](http://flowrepository.org/id/FR-FCM-Z244). We have performed gating cleanup on a subset of the dataset which can be dowloaded from zenodo. 
[Insert zenodo button]

## 2. Import data 

```{r download, eval=FALSE}

# Insert zenodo link

```

```{r sample setup, eval=FALSE}

```


```{r panel setup, eval=FALSE}

```


```{r import data, eval=FALSE}

```

Aternately you can download the imported RDS directly to skip the import step and save time:

```{r download rds, eval=TRUE}

# Insert zenodo RDS link

```


# Contact

For any queries relating to software:

* James W Opzoomer (james.opzoomer@kcl.ac.uk)
