---
title: "immunoCluster"
author: "James Opzoomer, Kevin Blighe, Jessica Timms"
date: "`r Sys.Date()`"
package: "`r packageVersion('immunoCluster')`"
output:
  github_document:
    toc: true
    toc_depth: 2
fig_width: 7
always_allow_html: true
---

## 1. Introduction to immunoCluster

The immunoCluster package uses the [scDataViz bioconductor package's](https://bioconductor.org/packages/devel/bioc/vignettes/scDataviz/inst/doc/scDataviz.html) vizualization tools and adaptation of the [SingleCellExperiment](https://www.bioconductor.org/packages/release/bioc/vignettes/SingleCellExperiment/inst/doc/intro.html) data structure to build simple and flexible cytometry analysis workflows like those outlined in Nowicka et al. (2017) [CyTOF workflow: differential discovery in high-throughput high-dimensional cytometry datasets](https://f1000research.com/articles/6-748). 

The package provides a broad toolkit to carry out immune profiling from both liquid and imaging high-dimensional mass (CyTOF) and flow cytometry data. immunoCluster features standardised data infrastructure, making use of the SingleCellExperiment class object, scDataViz interactive visualization tools along with convenient implementations for popular dimensionality reduction and clustering algorithms designed for a non-specialist. To learn using immunoCluster we have provided a walkthrough below exploring some of the package's basic functionality on a published dataset.  

An in depth report of immunoCluster's use to create liquid and imaging mass cytometry analysis pipelines, as well as its use for fluorescent flow cytometry analysis is outlined in our recent preprint [here](https://www.biorxiv.org/content/10.1101/2020.09.09.289033v1).  

## 2. Main functionalities of immunoCluster

The basic pipeline of immunoCluster analysis involves:

* Creating file and marker labels to generate the experimental design

* Importing .fcs data into the SingleCellExperiment object

* Exploratory analysis of the data

* Dimensionality reduction 

* Clustering and biological interpretation of the clusters

* Statistitcal testing for differences in cluster abundance and marker expression

```{r, echo = FALSE, message = FALSE}
  library(knitr)
  library(kableExtra)
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE, fig.align = 'center')
```

# Installation

## 1. Install the package from github

Install the development version of the package, installation will usually take a few minutes, depending on the number of dependencies that have been installed on your pc. Install immunoCluster from github here:

```{r getPackageDevel, eval = FALSE}
  # Install devtools for github installation if not present
  require(devtools)
  
  # Install the ImmunoCluster package
  devtools::install_github("kordastilab/ImmunoCluster")
  
  # May need to run this line if an R build version error while installing appears
  # Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")
```

## 2. Load required packages


Load necessary packages:
```{r Load, eval=TRUE, results="hide"}
  library(immunoCluster)
  library(stringr)
  library(cowplot)
  library(RColorBrewer)
```

# Walkthrough: Immune profiling PBMC from a two condition experimental design

## 1. Introduction

In this walkthrough we will learn to perform immune profiling with immunoCluster to examine changes in immune populations from human PBMC following bone marrow transplantation (BMT). 

We will start with pre-gated .fcs data files. The samples were taken from patients following hematopoetic reconstitution in leukemia patients following BMT. The dataset represents 15 individuals, sampled at two time points (Day 30 and 90) after BMT, for a total of 28 samples. Of these patients, a small subset suffered from graft versus host disease (GvHD, n = 3), whereas most other patients did not experience such complications. 

We have performed gating cleanup on a subset of the dataset which can be dowloaded from zenodo, along with all the other data required for the walkthough.

The data was originally reported in [Comprehensive Immune Monitoring of Clinical Trials to Advance Human Immunotherapy](https://www.sciencedirect.com/science/article/pii/S2211124719308228). The full raw data (not used here, but which was used in the immunoCluster paper) is available [here](http://flowrepository.org/id/FR-FCM-Z244).

All the data required can be downloaded from Zenodo. [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3801882.svg)](https://doi.org/10.5281/zenodo.3801882)


## 2. Download and import data 

You can download all the .fcs files from zenodo [here](https://doi.org/10.5281/zenodo.3801882), or alternatively you can download the SingleCellExperiment object directly into memory and skip the import process just below. The fcs files can be downloaded from this [download_link](https://zenodo.org/record/3801882/files/fcs_data.zip?download=1).

The experimental design informs the what metadata we will need to incorporate into the SingleCellExperiment (SCE) object, here we have created one table (from an excel file), that contains five columns, and one row for each .fcs file. We will include the .fcs filename, the GvHD "condition" of the patient,  patient_id, a unique sample_id, and the day_id representing the day following BMT that the sample was taken. [Download_link](https://zenodo.org/record/3801882/files/sample_metadata.xlsx?download=1)

```{r sample setup, eval=TRUE}

# Download sample metadata from Zenodo
sample_metadata_df = readRDS(url("https://zenodo.org/record/3801882/files/sample_metadata.rds"))

head(sample_metadata_df)
```

We also create a table that represents the marker panel information. This table contains a row for every channel in the original fcs file. The columns represent the original name of the channel and the channel name we want to rename it to in antigen. We then optionally include columns of binary 1 and 0s that allow us to group markers into relevant sets, here we have two sets, one group of markers for clustering and another for downstream expression analysis on our cell populations of interest. Any marker that is not included in either group will be discarded when we build the SCE object. This table must be an exact match to the order that the markers are represnted in the raw .fcs files or the channel renaming will be incorrect or fail. [Download_link](https://zenodo.org/record/3801882/files/panel_metadata.xlsx?download=1).

```{r panel setup, eval=TRUE}

# Download panel metadata from Zenodo
panel_metadata_df = readRDS(url("https://zenodo.org/record/3801882/files/panel_metadata.rds"))

head(panel_metadata_df)

# Create marker grouping designations
clustering_markers = panel_metadata_df$antigen[panel_metadata_df$clustering ==1]
activation_markers = panel_metadata_df$antigen[panel_metadata_df$activation == 1]
col_names = panel_metadata_df$antigen[panel_metadata_df$clustering == 1 | panel_metadata_df$activation == 1 ]

# Select columns to dicard from fcs file
discard = setdiff(panel_metadata_df$antigen, c(clustering_markers, activation_markers))
queries = str_detect(panel_metadata_df$antigen, paste(discard, collapse = "|"))

if(length(discard) > 0){colsDiscard = panel_metadata_df$metal[queries]
}else{colsDiscard = ''}

```

We next load the .fcs data and the metadata into the SCE object. Here the number of rows in the metadata must match the number of .fcs files to be analysed. We have the option to specify the scaling transformation and the cofactor. We have used asinh with cofactor 5. It is possible to downsaple from each file or by a specified metadata grouping, although here we have not downsampled in this case. The columns are also renamed as specified in the panel_metadata table and columns not needed are discarded.

```{r import data, eval=FALSE}

# Collect all .fcs files from a directory
filelist = list.files(
      path = "fcs_data/",
    pattern = "*.fcs|*.FCS",
    full.names = TRUE)

# Select which metadata to include from the sample metadata table
metadata = data.frame(
      file = filelist, 
      group = sample_metadata_df$sample_id,
      condition = sample_metadata_df$condition,
      patient_id = sample_metadata_df$patient_id,
      day_id = sample_metadata_df$day_id,
      row.names = filelist,
      stringsAsFactors = FALSE)

# Create SCE object with CyTOF data inside
require(SingleCellExperiment)
  sce_gvhd = processFCS(
    files = filelist,
    metadata = metadata,
    filter = FALSE, # Do not perform noise filtering
    transformation = TRUE, # Transform data
    transFun = function (x) asinh(x), # Using asinh function (cofactor 5 is used by default)
    downsampleVar = NULL, # No downsampling by variance
    downsample = NULL, # Downsample to n cells per downsample_grouping 
    downsample_grouping = NULL, # if downsample = n, downsample n cells by the specified metadata grouping.
    newColnames = col_names, # Rename columns from panel metadata 
    colsDiscard = colsDiscard) # Discard columns not selected for downstream analysis

sce_gvhd
```

Using the code chunk below you can download the imported r data object with the metatdata incorporated, directly into memory to skip the import steps to save time:

```{r download rds, eval=TRUE}
# Download complete ImmunoCluster SCE object from zenodo 
sce_gvhd = readRDS(url("https://zenodo.org/record/3801882/files/sce_gvhd.rds"))

sce_gvhd
```

## 2. Data Exploration

We provide several functions to examine global patterns within the data, they can be used to get an idea of the influence of conditions/treatment or batch effects within the data. The mdsplot() function will generate a multi-dimensional scaling (MDS) plot on median expression values for each channel. MDS plots represent an unsupervised way to indicate similarities between samples at a global level before more in depth analsysis. In the example here we can see that there is no observable high-level difference between the two GvHD condition samples, however there does appear to be a weak separation between samples collected at D30 and D90 post BMT. Here and throughout the pipeline, colkey can be used to specify plotting colors for graphing conditions. 

```{r mds_plot, fig.width=11, fig.height=3.75, eval=TRUE}
mds1 = mdsplot(sce_gvhd, 
               feature = "condition", 
               colkey = c(None = 'royalblue', GvHD = 'red2'))

mds2 = mdsplot(sce_gvhd, 
               feature = "day_id", 
               colkey = c(D30  = 'darkorange1', D90 = 'darkgreen'))

plot_grid(mds1, mds2,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 20)
```

Further to the mds plots,the medianHeatmap() function will generate a heatmap with median marker expression across all (default) or a selection (with markers = ). The heatmap is clustered over rows and columns. The heatmap can be generated on the asinh tranformed values (default) or 0-min to 1-max scaled (with scale_01 = T). The heatmap should provide insight as to which markers are highly expressed and whether their marker expression profile is associated with a metadata condition. Highly expressed markers will have a greater influence on clustering and this will help understand clustering results generated downstream.

```{r sample heatmap, eval=TRUE}
feature = medianHeatmap(sce_gvhd, grouping = "group", feature = "condition", feature_cols = c('royalblue', 'red2'))
```

## 3. Dimensionality reduction and clustering

immunoCluster supports multiple different dimensionality reduction algorithms and has UMAP and tSNE functionality implemented below with the ability to downsample before running the algorithms. We run both UMAP and tSNE on selected lineage markers, downsampling to 1000 randomly selected cells per sample id 'group' metadata slot (selected by specifying a downsample_grouping metadata slot) to reduce runtime. The dimensionality reductions are stored in the reducedDimNames slot and any other dimensionality reduction, like PCA, can also be stored in the SCE object in parallel.

```{r run UMAP, message=FALSE, eval=TRUE}
require(umap)
# Run UMAP and store in sce object
sce_gvhd = performUMAP(sce_gvhd, 
                       downsample = 1000, # Downsample to 1000 cells per downsample_grouping condition
                       downsample_grouping = "group", # Downsample by sample id 'group' metadata slot
                       useMarkers = clustering_markers) # Run UMAP only on selected markers

# Run tSNE and store in sce object
sce_gvhd = performTSNE(sce_gvhd, 
                       downsample = 1000, # Downsample to 1000 cells per downsample_grouping condition
                       downsample_grouping = "group",  # Downsample by sample id 'group' metadata slot
                       useMarkers = clustering_markers) # Run UMAP only on selected markers

sce_gvhd
```

The markerExpression() function will allow us to view the UMAP dimensionality reductions, overlaying marker expression of our major immune supopulation lineage markers as an exporatory analysis of the resultant UMAP. The dimensionality reduction data to use is specified with the reducedDim parameter.The markers and several ggplot style plotting parameters can be specified resulting in tiled plots of the UMAP with marker expression as colour. 

```{r, fig.width=18, fig.height=10, eval=TRUE}
expression_markers =  c('CD3', 'CD4', 'CD8a', 'CD11b', 'CD19', 'CD56')

exp_plot_umap = markerExpression(sce_gvhd,
    markers = expression_markers,
    reducedDim = 'UMAP',
    title = 'UMAP',
    nrow = 1, ncol = 6,
    pointSize = 0.05,
    legendKeyHeight = 1.0,
    legendLabSize = 14,
    stripLabSize = 20,
    axisLabSize = 18,
    titleLabSize = 20,
    captionLabSize = 22)

exp_plot_tsne = markerExpression(sce_gvhd,
    markers = expression_markers,
    reducedDim = 'Rtsne',
    title = 'tSNE',
    nrow = 1, ncol = 6,
    pointSize = 0.05,
    legendKeyHeight = 1.0,
    legendLabSize = 14,
    stripLabSize = 20,
    axisLabSize = 18,
    titleLabSize = 20,
    captionLabSize = 22)

plot_grid(exp_plot_umap, exp_plot_tsne,
    labels = c('A','B'),
    nrow = 2, align = "l", label_size = 24)
```

immunoCluster provides several wrapper functions to perform unsupervised clustering on your data. Currently users can implement either Rphenograph or FlowSOM clustering. Here we have used an ensemble method of FlowSOM and ConsensusCLusterPlus, which allows us to perform a fast, high resolution clustering on our dataset. We perform the clustering on all cells using a selection of lineage markers, with a final desired cluster number of up to 10, specified by the parameter k. FlowSOM will cluster cells into 100 SOM codes defined by the (dimensions of som_x and som_y) and these will be clustered into 2-k clusters and all of these clustering results are saved as metadata in the SCE object. 

```{r run clustering, message=TRUE, eval=TRUE}
sce_gvhd = runFlowSOM(sce_gvhd, k = 10, markers = clustering_markers, som_x = 10, som_y = 10)
```

The clustering identities are stored in the metadata dataframe. The per cell membership of the 100 SOM codes is stored under som_codes and the respective clusterings are stored under flowsom_cc_k(k=n). We can vizualise the clustering or any other metatdata on the dimnesionality plot using the function metadataPlot() (which maps to UMAP by default but can specify of dimension reductions that are stored like tSNE). We can also vizualise the abundance of each cluster as a proportion of total cells per sample using the plotAbundance() function. 

```{r viz flowSOM, fig.width=12, fig.height=6, eval=TRUE}
# Plot UMAP with flowsom k=10 clustering overlay
fsom_k10 = metadataPlot(sce_gvhd,
    colby = 'flowsom_cc_k10',
    title = '',
    reducedDim = 'UMAP',
    legendPosition = 'right',
    legendLabSize = 8,
    axisLabSize = 10,
    titleLabSize = 10,
    subtitleLabSize = 1,
    captionLabSize = 16)

# Specify all clusters to plot abundance for
fsom_clusters = unique(sce_gvhd@metadata$flowsom_cc_k10)

# Create boxplot of flowsom k=10 abundance 
fsom_abundance = plotAbundance(sce_gvhd,
              clusters = fsom_clusters,
              clusterAssign = 'flowsom_cc_k10',
              feature = NULL,
              legendPosition = 'none', 
              stripLabSize = 10,
              axisLabSize = 14,
              titleLabSize = 12,
              subtitleLabSize = 10,
              captionLabSize = 18)


plot_grid(fsom_k10, fsom_abundance,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 20, rel_widths = c(1.4, 1))
```

## 4. Biological annotation of the clustering

The medianHeatmap() function can also be used to display the median marker expression values of each flowSOM metacluster, by specifying the metadata clustering in the grouping parameter. Here the median values are 0-1 scaled to emphasize the differences between clusters (using scale_01 = T). 

```{r diagnostic heatmap, fig.width=5, fig.height=6, eval=TRUE}
# With feature
feature = medianHeatmap(sce_gvhd, grouping = "flowsom_cc_k10", feature = NULL, scale_01 = T)
```
 
 The heatmap can inform the supposed biological identity of the clusters generated by flowSOM and consesusClusterPlus. Another way to understand which markers are enriched per cluster is by using the finMarkers() function from the scran package. Here we perform a wilcox test to understand which markers have a LogFC > 1.5 enrichment between one cluster and several others.
 
```{r enriched markers, eval=TRUE}
library(scran)
library(tibble)

# Generate list of upregulated markers
cytof_markers = findMarkers(assay(sce_gvhd), sce_gvhd@metadata$flowsom_cc_k10, test.type="wilcox", direction="up",lfc=1.5, pval.type="some")

cluster_markers = NULL
top_n = 3 # The number of top upregulated markers to display
for (i in 1:length(cytof_markers)) {
  
  deg_markers = cytof_markers[[i]][1:top_n,1:2]
  
  deg_markers = rownames_to_column(as.data.frame(deg_markers), var = "marker")
  
  markers = data.frame(cluster = rep(i, top_n), deg_markers)
  
  cluster_markers = rbind(cluster_markers, markers)
  
}

# Show top 3 DEG markers per cluster
head(cluster_markers)
```

From this information we can assign identities to the clusters using the setClusterIdents() function. We feed a vector of original clusters and the new clusters we wish to assign these original cluster assignments to. The new cluster assignments will be stored in the metadata as cell_annotation. We can then assign this cell_annotation to a new metadata slot if we wish to store and reassign using setClusterIdents() again. 

```{r merge clusters, fig.width=11, fig.height=6, eval=TRUE}
# Assignments
original_cluster_ident = as.character(rep(1:10))
new_cluster_ident = c("Monocytes", "cDC", "NKT Cells", "CD4+ T Cells", "B Cells","gd T Cells", "CD8+ T Cells", "Basophils", "CD4+ Treg", "NK")

# Set the new annotations
sce_gvhd = setClusterIdents(sce_gvhd, orig.ident = original_cluster_ident , new.ident = new_cluster_ident, clustering = 'flowsom_cc_k10')

# Define population colours
cols = brewer.pal(10, "Set3")

# Assign to populations for plotting
col_key = c("Monocytes" = cols[1], "cDC" = cols[2], "NKT Cells" = cols[3], "CD4+ T Cells" = cols[4], "B Cells" = cols[5], "gd T Cells" = cols[6], "CD8+ T Cells" = cols[7], "Basophils" = cols[8], "CD4+ Treg" = cols[9], "NK" = cols[10])

# Vizualise in UMAP space
cell_annotation = metadataPlot(sce_gvhd,
    colby = 'cell_annotation',
    colkey = col_key, # Add colours for plotting 
    title = '',
    legendLabSize = 8,
    axisLabSize = 20,
    titleLabSize = 20,
    subtitleLabSize = 16,
    captionLabSize = 16)

annotated_clusters = unique(sce_gvhd@metadata$flowsom_cc_k10)

annotated_abundance = plotAbundance(sce_gvhd,
              clusters = annotated_clusters,
              clusterAssign = 'cell_annotation',
              feature = NULL,
              colkey = col_key,
              legendPosition = 'none', 
              stripLabSize = 10,
              axisLabSize = 14,
              titleLabSize = 12,
              subtitleLabSize = 10,
              captionLabSize = 18)

plot_grid(cell_annotation, annotated_abundance, 
    labels = c('A','B'), align = "l", label_size = 20, rel_widths = c(1.5,1), rel_heights = c(1.3,1))
```

The plotAbundance() function additionally allows to creat bar plots but specifying graph_type = "bar". Specfying a metadata feature to split the data by will allow to another level of comparison if wanted, first by plotting abundance by sample as a stacked bar and then arranging the samples by GvHD condition.  

```{r bar plot, fig.width=6, fig.height=4, eval=TRUE}
bar_plot = plotAbundance(sce_gvhd,
              graph_type = 'bar',
              clusters = annotated_clusters,
              clusterAssign = 'cell_annotation',
              feature = 'condition',
              colkey = col_key,
              legendLabSize = 7,
              stripLabSize = 22,
              axisLabSize = 22,
              titleLabSize = 22,
              subtitleLabSize = 18,
              captionLabSize = 18)

bar_plot
```

Once we have defined the biological identity of our clusters we can create a final heatmap to demonstrate that the marker expression patterns are consistent with our labels. Here we use medianHeatmap to create 0-1 normalised heatmap per cluster. We can also modify the the heat bar from the default greens to a black-white gradient (using heat_bar = "greys").

```{r final heatmap, fig.width=4, fig.height=6, eval=TRUE}
# With feature
feature = medianHeatmap(sce_gvhd, 
                        grouping = "cell_annotation",
                        markers = clustering_markers,
                        feature = NULL, 
                        scale_01 = T, 
                        heat_bar = "Greys")
```

## 5. Statistical testing for cluster differences in abundance and marker expression

We can now visualise the data by disease status to better understand the differences that may help us define biomarkers to predict GvHD status post BMT. We can also use the metadataPlot() function to split the UMAP by GvHD status by defining colkey and splitting the UMAP plot by condition with facet_wrap(~condition). Also by adding a feature parameter to the plotAbundance function we can split the boxplot by GvHD status allowing us to investigate changes changes in subset abundance by disease status or indeed, any other metadata feature slot.  

```{r cell annotations, fig.width=10, fig.height=5, eval=TRUE}
cell_annotation_dif = metadataPlot(sce_gvhd,
    colby = 'condition',
    title = '',
    colkey = c(None = 'royalblue', GvHD = 'red2'),
    legendPosition = 'right',
    pointSize = 0.3,
    legendLabSize = 8,
    axisLabSize = 14,
    titleLabSize = 16,
    subtitleLabSize = 16,
    captionLabSize = 16)

cell_annotation_dif = cell_annotation_dif + facet_wrap(~condition)

abundance_dif = plotAbundance(sce_gvhd,
              clusters = annotated_clusters,
              clusterAssign = 'cell_annotation',
              feature = 'condition',
              colkey = c(None = 'royalblue', GvHD = 'red2'),
              legendPosition = 'none', 
              stripLabSize = 10,
              axisLabSize = 14,
              titleLabSize = 12,
              subtitleLabSize = 10,
              captionLabSize = 18)

plot_grid(cell_annotation_dif, abundance_dif,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 20, rel_widths = c(1.6,1))
```

Finally we can perform a wilcoxon rank sum test to see if we can detect statistically significant changes between our two experimental conditions. stat_test_clust_results() will perform the wilcoxon test on our data and produce a results table along with Log fold change and the proportion of each sample that each cluster represents. We can thes use volcanoPlot() to generate a volcano plot, which shows us that there is a significant difference between the abundance of B cells between our two conditions. We can see that in our dataset it appears that B Cells are markedly reduced in those patients that go on to develop GvHD.

```{r abundance test, fig.width=10, fig.height=5, eval=TRUE}
# This need the individual values
# This also need the right contrasts
stat_test_clust_results = stat_test_clust(sce_gvhd,
                        group = 'group', # group the files by sample ID metadata slot 'group'
                        clustering = 'cell_annotation', # The clustering label to use
                        feature = 'condition')

head(stat_test_clust_results[,c(1,14:18)])

gg_volcano = volcanoPlot(stat_test_clust_results, p_val = "padj", threshold = 0.12)

abundance_bcell = plotAbundance(sce_gvhd,
              clusters = c("B Cells", "Basophils"),
              clusterAssign = 'cell_annotation',
              feature = 'condition',
              colkey = c(None = 'royalblue', GvHD = 'red2'),
              legendPosition = 'right', 
              stripLabSize = 10,
              axisLabSize = 14,
              titleLabSize = 12,
              subtitleLabSize = 10,
              captionLabSize = 18)

plot_grid(gg_volcano, abundance_bcell,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 20, rel_widths = c(1,1.4))
```

We can perform the same process of applying a wilcoxon test for expression levels of markers, across all markers on all populations by using the stat_test_expression() function. We can plot the results a volcano plot as before using volcanoPlot(). Beside this can vizualise this cluster specific expression of certain makers with markerExpressionPerSample() to look at the median marker expression per sample on these clusters (or all clusters).

```{r marker test, fig.width=12, fig.height=7, eval=TRUE}
# Need to catch errors - cluster not there causes crash
marker_test = stat_test_expression(sce_gvhd,
                assay = 'scaled', # The sce assay slot
                grouping = 'group', # group the files by sample ID metadata slot 'group'
                feature = 'condition', # The contrast metatdata
                clusterAssign = 'cell_annotation') # The clustering label to use

sig_markers = marker_test[which(marker_test$p_val < 0.05),]

head(sig_markers)

# Create volcano plot
gg_volcano = volcanoPlot(marker_test, p_val = "p_val", threshold = 0.05)

# Difference in activation markers per cluster
pd_expression = markerExpressionPerSample(sce_gvhd,
    caption = '',
    clusters = c("CD8+ T Cells", "cDC"),
    feature = 'condition',
    clusterAssign = 'cell_annotation',
    markers = activation_markers[3:5],
    colkey = c('royalblue', 'red2'),
    title = '',
    stripLabSize = 12,
    axisLabSize = 10,
    titleLabSize = 18,
    subtitleLabSize = 10,
    captionLabSize = 10)


plot_grid(gg_volcano, pd_expression,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 20, rel_widths = c(1,1.4))

```

# Subsetting the SCE object

The SCE object can be subsetted for further clustering, for instance, in a situation when increased cluster resolution is desired or if in a situiation where you want to diplay the proportion of a certain cluster as a percentage of another parent populations, for instance CD4+ Tregs as a proportion of all CD4+ T cells. Subsetting can be performed using the subsetSCE() function. The function can subset cells on any specified metadata slot using dplyr syntax conditional statements. An couple of examples are presented below:

```{r subsetting_1}
# Subset by patient_ID
sce_subset = subsetSCE(sce_gvhd, patient_id %in% c("P1", "P9", "P15"))

table(sce_subset@metadata$patient_id)
```

```{r subsetting_2}
# Subset by clustering idenity
sce_cd4 = subsetSCE(sce_gvhd, cell_annotation %in% c("CD4+ T Cells", "Basophils"))

table(sce_cd4@metadata$cell_annotation)
```


# Session info
```{r session info, eval=TRUE}
sessionInfo()
```

# Contact

For any queries relating to software:

* Jessica Timms (jessica.timms@kcl.ac.uk)
